3D Magnetic modelling of ellipsoidal bodies

### 2017-02-14 manuscript/manuscript.tex
include refs about anisotropy

Based on the mathematical theory of the magnetic 
induction developed by \citet{poisson1824}, 
\citet{maxwell1873} affirmed that, if $U$ is the 
gravitational potential produced by any body 
with uniform density $\rho$ and arbitrary shape at 
a point $(x, y, z)$, then $-\frac{\partial U}{\partial x}$
is the magnetic scalar potential produced 
at the same point by the same body 
if it has a uniform magnetization oriented along $x$ 
with intensity $\rho$.
\citet{maxwell1873} generalized this idea as a way
of determining the magnetic scalar potential produced
by any body uniformly magnetized in a given direction.
By presuming that this uniform magnetization is due to
induction, he postulated that the resulting magnetic
field (intensity) at all points within the body must
also be uniform and parallel to the magnetization.
As a consequence, the gravitational potential 
$U$ at points within the body must be a quadratic 
function of the spatial coordinates.
Apparently, \citet{maxwell1873} was the first one to
postulate that ellipsoids are the only finite bodies 
having a gravitational potential which satisfies this 
property and hence can be uniformly magnetized 
in the presence of a uniform magnetic field.
This property can be extended to other bodies
defined as limiting cases of an ellipsoid
(e.g., spheres, elliptic cylinders), however all the
remaining non-ellipsoidal bodies cannot be
uniformly magnetized in the presence of a uniform
inducing field.

Self-demagnetization demagnetization gives rise to shape 
anisotropy.

Self-demagnetization is a function of the shape of 
the body.

The self-demagnetization contributes to decrease the
magnitude of the magnetization along the shortest 
dimensions of a body.

Ellipsoids are the only bodies which enable an
analytical computation of its self-demagnetization.

Ellipsoids can assume a wide range of shapes and
therefore can be used to represent different geological
bodies.

It is well-known by that the self-demagnetization can 
be neglected if the body has a susceptibility lower 
than 0.1 SI \citep{austin2014, clark2014, clark1986, 
emerson1985, eskola1980, guo1998, guo2001, purss2005,
hillan2013}.
However, neglecting the self-demagnetization in
geological bodies with high susceptibilities (> 0.1 SI) 
may strongly mislead the interpretation obtained from
magnetic methods.
\citet{farrar1979} demonstrated the importance of the
ellipsoidal model in taking into account the 
self-demagnetization and determining reliable 
drilling directions on the Tennant Creek field, 
Australia.
Posteriorly, \citet{hoschke1991} also showed how the
ellipsoidal model proved to be highly successful in
locating and defining ironstone bodies in the
Tennant Creek field.
\citet{clark2000} provides a good discussion about the
influence of the self-demagnetization in magnetic
interpretation of the Osborne copper-gold deposit, 
Australia. This deposit is hosted by ironstone 
bodies that have very high susceptibility.
According to \citet{clark2000}, neglecting the effects 
of self-demagnetisation led errors of $\approx 55^{\circ}$
in the interpreted dip. 
Recently, \citet{austin2014} used magnetic modelling 
and rock property measurements to show that,
contrary to previous interpretations, the magnetization
of the Candelaria iron oxide copper-gold deposit, 
Chile, is not dominated by the induced component.
Rather, the deposit has a relatively weak remanent 
magnetization and is strongly affected by 
self-demagnetization.
These examples show the importance of the ellipsoidal
model in producing trustworthy geological models of
high-susceptibility orebodies, which may save significant 
cost associated with drilling.

As we can see, a vast literature about the magnetic
modelling of ellipsoidal bodies was developed in which 
are to be found the names of many brilliant researchers.
Nevertheless, interest in this subject has not yet died
out, as is evidenced by the list of modern papers in this
field.
Besides, the geoscientific community lacks an easy-to-use 
tool to simulate the magnetic field produced by uniformly 
magnetized ellipsoids.
Such a tool could prove to be useful either for teaching
and researching geophysics.

To fill this gap, we provide here a set of routines to 
model the magnetic field produced by ellipsoids.
The routines are written in Python language as part of
the Fatiando a Terra \citep{uieda2013}, which is an 
open-source library for modelling and inversion in geophysics.
We attempt to use the best practices of continuous 
integration, documentation, unit-testing, and version-control
for the purpose of providing a reliable and easy-to-use
code.
It is hoped that the presentation of the theoretical and
practical aspects given here will be found useful to the
wide community of geoscientists.

@@

Let $(x, y, z)$ be a point referred to a Cartesian coordinate system with axes
$x$, $y$ and $z$ pointing to, respectively, North, East and down.
For convenience, we denominate this coordinate system as 
\textit{main coordinate system}.
Let us consider an ellipsoidal body with centre at the point $(x_{c}, y_{c}, z_{c})$,
semi-axes defined by positive constants $a$, $b$, $c$, 
where $a > b > c$, and orientation defined by
three angles $\alpha$, $\beta$, and $\gamma$.
The points $(x, y, z)$ located on the surface of this ellipsoidal body
satisfy the following equation:
\begin{equation}
(\mathbf{r} - \mathbf{r}_c)^T \mathbf{A} (\mathbf{r} - \mathbf{r}_c) = 1 \: ,
\label{eq:ellipsoid_surface}
\end{equation}
where $\mathbf{r} = [\begin{array}{ccc} x & y & z \end{array} ]^{\top}$,
$\mathbf{r}_{c} = [\begin{array}{ccc} x_{c} & y_{c} & z_{c} \end{array} ]^{\top}$,
$\mathbf{A}$ is a positive definite matrix given by
\begin{equation}
\mathbf{A} = \mathbf{V}
\left[ \begin{array}{ccc}
a^{-2} & 0 & 0 \\
0 & b^{-2} & 0 \\
0 & 0 & c^{-2} 
\end{array} \right] \mathbf{V}^{\top} \: ,
\label{eq:A}
\end{equation}
and $\mathbf{V}$ is an orthogonal matrix whose columns are defined
by unit vectors $\mathbf{v}_{1}$, $\mathbf{v}_{2}$, and $\mathbf{v}_{3}$.

>>>>

Based on the mathematical theory of the magnetic 
induction developed by \citet{poisson1824}, 
\citet{maxwell1873} affirmed that, if $U$ is the 
gravitational potential produced by any body 
with uniform density $\rho$ and arbitrary shape at 
a point $(x, y, z)$, then $-\frac{\partial U}{\partial x}$
is the magnetic scalar potential produced 
at the same point by the same body 
if it has a uniform magnetization oriented along $x$ 
with intensity $\rho$.
\citet{maxwell1873} generalized this idea as a way
of determining the magnetic scalar potential produced
by any body uniformly magnetized in a given direction.
By presuming that this uniform magnetization is due to
induction and that it is proportional to the resulting
magnetic field (intensity) inside the body, he postulated 
that the resulting field must also be uniform and 
parallel to the magnetization. This uniformity is due to
the fact that the resulting field is defined as the
negative gradient of the magnetic scalar potential
within the body.
As a consequence of this uniformity, 
the gravitational potential $U$ at points within the 
body must be a quadratic 
function of the spatial coordinates.
Apparently, \citet{maxwell1873} was the first one to
postulate that ellipsoids are the only finite bodies 
having a gravitational potential which satisfies this 
property and hence can be uniformly magnetized 
in the presence of a uniform magnetic field.
This property can be extended to other bodies
defined as limiting cases of an ellipsoid
(e.g., spheres, elliptic cylinders), however all the
remaining non-ellipsoidal bodies cannot be
uniformly magnetized in the presence of a uniform
inducing field.

Another particularity of ellipsoids are that they are
the only bodies which enable an analytical computation 
of its self-demagnetization.
The self-demagnetization contributes to decrease the
magnitude of the magnetization along the shortest 
axes of a body.
It is a function of the shape of the body and gives rise
to shape anisotropy \citep{thompson1986, clark1999, tauxe2003}.
It is well-known by that the self-demagnetization can 
be neglected if the body has a susceptibility lower 
than 0.1 SI \citep{austin2014, clark2014, clark1986, 
emerson1985, eskola1980, guo1998, guo2001, purss2005,
hillan2013}.
However, neglecting the self-demagnetization in
geological bodies with high susceptibilities (> 0.1 SI) 
may strongly mislead the interpretation obtained from
magnetic methods.
\citet{farrar1979} demonstrated the importance of the
ellipsoidal model in taking into account the 
self-demagnetization and determining reliable 
drilling directions on the Tennant Creek field, 
Australia.
Posteriorly, \citet{hoschke1991} also showed how the
ellipsoidal model proved to be highly successful in
locating and defining ironstone bodies in the
Tennant Creek field.
\citet{clark2000} provides a good discussion about the
influence of the self-demagnetization in magnetic
interpretation of the Osborne copper-gold deposit, 
Australia. This deposit is hosted by ironstone 
bodies that have very high susceptibility.
According to \citet{clark2000}, neglecting the effects 
of self-demagnetisation led errors of $\approx 55^{\circ}$
in the interpreted dip. 
Recently, \citet{austin2014} used magnetic modelling 
and rock property measurements to show that,
contrary to previous interpretations, the magnetization
of the Candelaria iron oxide copper-gold deposit, 
Chile, is not dominated by the induced component.
Rather, the deposit has a relatively weak remanent 
magnetization and is strongly affected by 
self-demagnetization.
These examples show the importance of the ellipsoidal
model in producing trustworthy geological models of
high-susceptibility orebodies, which may save significant 
cost associated with drilling.

A vast literature about the magnetic
modelling of ellipsoidal bodies was developed in which 
are to be found the names of many brilliant researchers.
Nevertheless, interest in this subject has not yet died
out, as is evidenced by a list of modern papers in this
field.
Besides, the geoscientific community lacks an easy-to-use 
tool to simulate the magnetic field produced by uniformly 
magnetized ellipsoids.
Such a tool could prove to be useful either for teaching
and researching geophysics.
To fill this gap, we provide here a set of routines to 
model the magnetic field produced by ellipsoids.
The routines are written in Python language as part of
the Fatiando a Terra \citep{uieda2013}, which is an 
open-source library for modelling and inversion in geophysics.
We attempt to use the best practices of continuous 
integration, documentation, unit-testing, and version-control
for the purpose of providing a reliable and easy-to-use
code.
It is hoped that the presentation of the theoretical and
practical aspects given here will be found useful to the
wide community of geoscientists.

@@

Let $(x, y, z)$ be a point referred to a Cartesian coordinate system with axes
$x$, $y$ and $z$ pointing to, respectively, North, East and down.
For convenience, we denominate this coordinate system as 
\textit{main coordinate system}.
Let us consider an ellipsoidal body with centre at the point $(x_{c}, y_{c}, z_{c})$,
semi-axes defined by positive constants $a$, $b$, $c$, 
where $a > b > c$, and orientation defined by
three angles $\alpha$, $\beta$, and $\gamma$.

DESCREVER A ORIENTACAO EM FUNCAO DOS ANGULOS DE GEOLOGIA ESTRUTURAL
\citep{clark1986, allmendinger2012}

The points $(x, y, z)$ located on the surface of this ellipsoidal body
satisfy the following equation:
\begin{equation}
(\mathbf{r} - \mathbf{r}_c)^T \mathbf{A} (\mathbf{r} - \mathbf{r}_c) = 1 \: ,
\label{eq:ellipsoid_surface}
\end{equation}
where $\mathbf{r} = [\begin{array}{ccc} x & y & z \end{array} ]^{\top}$,
$\mathbf{r}_{c} = [\begin{array}{ccc} x_{c} & y_{c} & z_{c} \end{array} ]^{\top}$,
$\mathbf{A}$ is a positive definite matrix given by
\begin{equation}
\mathbf{A} = \mathbf{V}
\left[ \begin{array}{ccc}
a^{-2} & 0 & 0 \\
0 & b^{-2} & 0 \\
0 & 0 & c^{-2} 
\end{array} \right] \mathbf{V}^{\top} \: ,
\label{eq:A}
\end{equation}
and $\mathbf{V}$ is an orthogonal matrix whose columns are defined
by unit vectors $\mathbf{v}_{1}$, $\mathbf{v}_{2}$, and $\mathbf{v}_{3}$.
